---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

##### Load the activity data file
```{r}
data <- read.csv("activity.csv", header=TRUE, sep = ",")
```

##### Convert the "date" values to date format
```{r}
data$date <- as.Date(data$date, "%Y-%m-%d")
```

##### Show the first few rows of the dataset

```{r}
head(data)
```

```{r}
str(data)
```

## What is mean total number of steps taken per day?
##### Calculate the total number of steps taken per day
```{r, echo=TRUE}
aggStepsDateSum <- aggregate(steps ~ date, data, sum, na.rm=TRUE)
names(aggStepsDateSum) <- c("date", "totalSteps")
```

##### Make a histogram of the total number of steps taken each day

```{r}
hist(aggStepsDateSum$totalSteps, breaks=20, col="green", 
     main="Total number of steps taken each day", 
     xlab="Total number of steps")
```

##### The mean of the total number of steps taken per day

```{r}
mean(aggStepsDateSum$totalSteps)
```

##### The median of the total number of steps taken per day

```{r}
median(aggStepsDateSum$totalSteps)
```

## What is the average daily activity pattern?

##### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
aggStepsIntervalMean <- aggregate(steps ~ interval, data, FUN=mean, na.rm=TRUE)
names(aggStepsIntervalMean) <- c("interval", "meanSteps")

head(aggStepsIntervalMean)
```

```{r}
plot(aggStepsIntervalMean, type = "l", main="Average number of steps",
     xlab="Consecutive 5-minute interval", ylab="Avg no. of steps taken")
```

##### Find which 5-minute interval contains the maximum number of steps

```{r}
maxInterval <- which.max(aggStepsIntervalMean$meanSteps)
aggStepsIntervalMean$interval[maxInterval]
```

## Imputing missing values

##### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missingValues <- sum(is.na(data))
missingValues
```

##### Imputing missing values by using the mean for a given interval

```{r}
# copy the dataframe to imputedData
imputedData <- data

# Fills in empty step data with mean for a given interval.
for (row in which(is.na(imputedData$steps))) {
        imputedData$steps[row] <-
           as.numeric(aggStepsIntervalMean[aggStepsIntervalMean$interval ==
                                           imputedData$interval[row],]["meanSteps"])
}
```

##### Show the first few rows of imputed data. Notice that the NA values for 2012-10-01 have been replaced.
```{r}
head(imputedData)
```

##### Check if there are anymore NA values
```{r}
sum(is.na(imputedData))
```

##### Make a histogram of the total number of steps taken each day.

```{r}
aggStepDate2 <- aggregate(steps ~ date, data=imputedData, FUN=sum)
names(aggStepDate2) <- c("date", "totalSteps")

# make the histogram
hist(aggStepDate2$totalSteps, breaks=20, col="blue", main="Histogram of total number of steps per day", xlab="[Imputed] Total number of steps in a day")
```

##### Calculate and report the mean and median again. 

```{r}
mean(aggStepDate2$totalSteps)
```

##### The median of the total number of steps taken per day. 

```{r}
median(aggStepDate2$totalSteps)
```

##### Note the change in the median value compared to the value calculated before filling in the missing values, but the mean value remains the same.

## Are there differences in activity patterns between weekdays and weekends?

##### Create a new factor variable "dayType" in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
imputedData$dayType <- as.factor(ifelse(weekdays(imputedData$date) %in%
                                    c("Saturday","Sunday"), "Weekend",
                                                            "Weekday")) 

aggStepsInt <- aggregate(steps ~ interval + dayType, data=imputedData, FUN=mean)
```

##### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
library(lattice)
xyplot(steps ~ interval | dayType, aggStepsInt, type = "l", layout = c(1, 2), 
       xlab="consecutive 5-minute interval", ylab="Avg no. of steps")
```

##### As you can see, the stepping activity is more uniformly distributed during weekends. Whereas you can see the spike at around 800th minute during weekdays.